Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-09 infercnv/reports/gains_and_losses.Rmd
use dittoseq for plotting umaps

First, we load the CITE-seq object. We convert it to a SingleCellExperiment.

@@

}, file = 'sce.rds')

sce <- xfun::cache_rds({

@@

Next, we add the CNV metadata from the inferCNV run to the UMAP embeddings. We must wrangle it from wide- to long-format.

```{r infercnv_output}
cnvs <- xfun::cache_rds({

  # get cnv metadata
  cnvs <- 
    list.files(
      infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(infercnv_dir, file) %>% 
        read.table(sep = '\t') %>% 
        tibble::as_tibble(rownames = 'cell')
      }) %>%
    dplyr::bind_rows(.id = 'id') %>%
    # convert from wide to long format
    tidyr::pivot_longer(
      cols = tidyselect::starts_with(c('has_', 'proportion_', 'top_'))
    )

  # get chr / event / name
  names <-
    cnvs %>%
    dplyr::distinct(name) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      chr = gsub('.*_', '', name),
      event = gsub(paste0('_', chr), '', name) %>% sub('.*_', '', .),
      measure = sub(paste0('_', event, '_', chr), '', name),
      chr = dplyr::case_when(
        measure == 'top' ~ paste0('chr', chr), TRUE ~ chr) %>% 
        {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())}) %>%
    dplyr::ungroup()

  # combine umap, names, and cnvs
  cnvs <-
    cnvs %>%
    # add names
    dplyr::left_join(names, by = 'name')

  # return
  cnvs

}, file = 'cnvs.rds')
```

@@

Now, we will plot gain and loss events by chromosome

```{r plot_gains_and_losses, fig.height = 20}
# function: plot gains and losses
plot_gains_and_losses <- 
  function(cnvs, annot) {
    cnvs %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'proportion', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(proportion = mean(value)) %>%
      ggplot(aes(x = chr, y = id, alpha = proportion, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines"),
            axis.text.y = element_text(size = 3),
            axis.text.x = element_text(angle = 270, hjust = 0, vjust = 0.5)) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(annotation ~ event)
  }

@@

We are most interested in the chr7 gain events in fibroblasts.

```{r chr7_prop_dist}
p_dat <-
  cnvs %>%
  dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
  dplyr::group_by(id, cluster_name, cell_type) %>%
  dplyr::mutate(median_proportion = median(value)) %>%
  dplyr::ungroup()

p_dat %>%
  ggplot(aes(x = value)) + 
  geom_density() +
  geom_vline(data = p_dat, aes(xintercept = median_proportion)) +
  ggh4x::facet_nested(~ subcluster + cell_type)

plot_prop_chr7_gain <- function(p_dat, position_arg) {
  p_dat %>%
    ggplot() +
    geom_bar(aes(x = cell_type, fill = value, group = value),
             position = position_arg) +
    theme_classic() +
    scale_fill_viridis_b(n.breaks = 9) 
}

plot_prop_chr7_gain(p_dat, position_stack(reverse = T))
plot_prop_chr7_gain(p_dat, position_fill(reverse = T))
p_dat %>%
  {split(., .$id)} %>%
  purrr::map2(names(.), ., function(id, df) {
    plot_prop_chr7_gain(df, position_fill(reverse = T)) +
      ggtitle(paste0('chr7 gain count - ', id))
    plot_prop_chr7_gain(df, position_stack(reverse = T)) +
      ggtitle(paste0('chr7 gain proportion - ', id))
  })
```

@@

```{r pseudobulk}
pseudobulk <- xfun::cache_rds({

  # pseudobulk the sce object by chr7 status
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce, 
      group_by = vars()
    )

>>>>

First, we load the CITE-seq object. We convert it to a SingleCellExperiment object.

@@

}, file = 'sce.rds')
```

Next, we add the CNV metadata from the inferCNV run to the object.

```{r get_infercnv}
sce <- xfun::cache_rds({

@@

@@

First, we plot UMAPs by ID and celltype.

```{r plot_umaps, fig.height = 10}
dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
  coord_fixed()
```

```{r plot_umaps_by_id, fig.height = 30}
dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2, 
                       split.by = 'sample') +
  coord_fixed()
```

Then, we visualise chr7 and chr5 gains.

```{r plot_chr_gains}
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr7', size = 0.2) +
  coord_fixed()
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr7', size = 0.2, split.by = 'sample') +
  coord_fixed()
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr5', size = 0.2) +
  coord_fixed()
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr5', size = 0.2, split.by = 'sample') +
  coord_fixed()
```

Now, we will plot gain and loss events by chromosome.

```{r plot_gains_and_losses, fig.height = 20}
# mean proportion loss / gain per chromosome per celltype per id
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::group_by(sample, cell_type) %>%
  dplyr::summarise(
    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
  dplyr::ungroup()

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::set_names() %>%
  purrr::map(function(direction) {
    list(
      matrix = 
        p_dat %>% 
        dplyr::select(dplyr::contains(paste0('proportion_', direction))) %>%
        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1'))) %>%
        as.matrix(),
      annot = p_dat %>% dplyr::select(sample, cell_type)
    )
  })

# plot heatmaps
pheatmap::pheatmap(p_dat$dupli$matrix, cluster_cols = F, cluster_rows = F)
pheatmap::pheatmap(sce@colData[,grepl('proportion_dupli', colnames(sce@colData))])

# function: plot gains and losses
plot_gains_and_losses <- 
  function(cnvs, annot) {
    cnvs %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'proportion', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(proportion = mean(value)) %>%
      ggplot(aes(x = chr, y = id, alpha = proportion, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines"),
            axis.text.y = element_text(size = 3),
            axis.text.x = element_text(angle = 270, hjust = 0, vjust = 0.5)) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(annotation ~ event)
  }

@@

We are most interested in the chr7 gain events in fibroblasts. We visualise their distribution across the dataset.

```{r chr7_prop_dist}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::filter(!is.na(proportion_dupli_chr7))

plot_prop_chr7_gain <- function(p_dat, position_arg) {
  p_dat %>%
    ggplot() +
    geom_bar(aes(x = cell_type, 
                 fill = proportion_dupli_chr7, 
                 group = proportion_dupli_chr7),
             position = position_arg) +
    scale_x_discrete(guide = guide_axis(angle = -45)) +
    theme_classic() +
    scale_fill_viridis_b(n.breaks = 9) 
}

plot_prop_chr7_gain(p_dat, position_stack(reverse = T)) +
  ggtitle('chr7 gain count - all cells')
plot_prop_chr7_gain(p_dat, position_fill(reverse = T)) +
  ggtitle('chr7 gain proportion - all cells')
p_dat %>%
  {split(., .$id)} %>%
  purrr::map2(names(.), ., function(id, df) {
    plot_prop_chr7_gain(df, position_fill(reverse = T)) +
      ggtitle(paste0('chr7 gain count - ', id))
    plot_prop_chr7_gain(df, position_stack(reverse = T)) +
      ggtitle(paste0('chr7 gain proportion - ', id))
  })
```

@@

```{r pseudobulk_chr7_gain}
pseudobulk <- xfun::cache_rds({

  # we assign chr7 status as >90% of genes affected by a duplication
  sce@colData$chr7_gain <-
    as.numeric(sce@colData$proportion_dupli_chr7 > 0.9)

  # pseudobulk the sce object by chr7 status
  reduced_sce <-
    glmGamPoi::pseudobulk_sce(
      sce, 
      group_by = vars(chr7_gain, cell_type)
    )
